Reanalysis Step 4: Outcome Analysis with Three-Way Interaction

### 2025-08-20 reanalysis-4-outcome_analysis.qmd
calc df & reformat

# Additional packages
library(emmeans)
library(margins)
library(survey)
library(ggplot2)
library(gridExtra)
library(knitr)
library(tidyr)
library(dplyr)
library(mitools)  # For pooling multiple imputation results
library(mice)

@@

  if (length(threeway_terms) > 0) {
    message("  - Three-way interaction terms found: ", length(threeway_terms))
    for (term in threeway_terms) {
      p_val <- coef_summary[term, "Pr(>|t|)"]
      message("    ", term, ": p = ", round(p_val, 4))
    }
  } else {
    message("  - Warning: No three-way interaction terms found")
  }

@@

message("All models fitted successfully")

@@

# Remove any NULL entries if they exist
models <- models[!sapply(models, is.null)]
if (length(models) == 0) {
  stop("No valid models to pool")
}

message("Ready to pool ", length(models), " models")
```

@@

```{r pool-results}
# Use mitools for robust pooling implementation
# MIcombine properly implements Rubin's rules with Barnard-Rubin adjustment

# Ensure models is a proper list for MIcombine
if (!is.list(models)) {
  stop("Models must be a list")
}

@@

pooled_model <- MIcombine(models)

# Extract pooled results
pooled_summary <- summary(pooled_model)

# Get pooled variance-covariance matrix for later use
pooled_vcov <- vcov(pooled_model)

pooled_results <- data.frame(
  Term = rownames(pooled_summary),
  Estimate = pooled_summary[, "results"],
  SE = pooled_summary[, "se"],
  statistic = pooled_summary[, "(lower"],  
  df = pooled_summary[, "df"],
  p.value = pooled_summary[, "upper)"],
  CI_Lower = pooled_summary[, "2.5 %"],
  CI_Upper = pooled_summary[, "97.5 %"],
  OR = exp(pooled_summary[, "results"]),
  OR_CI_Lower = exp(pooled_summary[, "2.5 %"]),
  OR_CI_Upper = exp(pooled_summary[, "97.5 %"])
)

# Clean column names if needed
names(pooled_results) <- c("Term", "Estimate", "SE", "t_stat", "df", "P_Value", 
                           "CI_Lower", "CI_Upper", "OR", "OR_CI_Lower", "OR_CI_Upper")
# Ensure proper formatting
rownames(pooled_results) <- NULL

@@

```{r pooling-diagnostics}
pool_diag <- data.frame(
  Metric = c("Number of imputations", 
             "Number of parameters",
             "Average degrees of freedom"),
  Value = c(n_imputations,
            nrow(pooled_results),
            round(mean(pooled_results$df), 1))
)

@@

key_terms <- grep("ever_lapse", pooled_results$Term, value = TRUE)[1:min(3, length(grep("ever_lapse", pooled_results$Term, value = TRUE)))]
key_df <- pooled_results[pooled_results$Term %in% key_terms, c("Term", "df")]
key_df$Term <- substr(key_df$Term, 1, 40)
key_df$df <- round(key_df$df, 1)

kable(key_df, caption = "Degrees of freedom for key terms")

@@

message("Performing pooled joint hypothesis tests...")

@@

if (length(threeway_terms_test) > 0) {
  hypothesis_3way <- paste0(threeway_terms_test, " = 0")

  # Pool test across imputations
  pooled_test_3way <- pool_chi_square_tests(models, hypothesis_3way, "Three-way interaction")

  message("1. Three-way interaction test (H0: all three-way terms = 0):")
  message("   Chi-sq = ", round(pooled_test_3way$D_bar, 3))
  message("   df = ", pooled_test_3way$df)
  message("   p-value = ", format.pval(pooled_test_3way$p_value, digits = 4, eps = 0.0001))

  if(pooled_test_3way$p_value < 0.05) {
    message("   Interpretation: The effect of lapsing varies by BOTH DR severity AND treatment status")
  } else {
    message("   Interpretation: No significant three-way interaction detected")
  }
}

@@

if (length(lapse_dr_terms) > 0) {
  hypothesis_lapse_dr <- paste0(lapse_dr_terms, " = 0")

  # Pool test across imputations
  pooled_test_lapse_dr <- pool_chi_square_tests(models, hypothesis_lapse_dr, "Lapse × DR interaction")

  message("2. Lapse × DR severity interaction test:")
  message("   Chi-sq = ", round(pooled_test_lapse_dr$D_bar, 3))
  message("   df = ", pooled_test_lapse_dr$df)
  message("   p-value = ", format.pval(pooled_test_lapse_dr$p_value, digits = 4, eps = 0.0001))
}

@@

if (length(lapse_tx_terms) > 0) {
  hypothesis_lapse_tx <- paste0(lapse_tx_terms, " = 0")

  # Pool test across imputations
  pooled_test_lapse_tx <- pool_chi_square_tests(models, hypothesis_lapse_tx, "Lapse × treatment interaction")

  message("3. Lapse × treatment interaction test:")
  message("   Chi-sq = ", round(pooled_test_lapse_tx$D_bar, 3))
  message("   df = ", pooled_test_lapse_tx$df)
  message("   p-value = ", format.pval(pooled_test_lapse_tx$p_value, digits = 4, eps = 0.0001))
}
```

@@

  message("Calculating marginal means and contrasts across all imputations...")
  message("This may take a few moments...")

@@

  for (i in 1:n_imputations) {
    message("  Processing imputation ", i, "/", n_imputations)

>>>>

# Additional packages
library(emmeans)
library(margins)
library(survey)
library(ggplot2)
library(gridExtra)
library(knitr)
library(tidyr)
library(dplyr)
library(mice)  # For pool() function with Barnard-Rubin df adjustment
library(mitools)  # For additional MI utilities if needed

@@

  # Track progress
  model_progress <- rbind(model_progress, data.frame(
    Imputation = i,
    Converged = model$converged,
    ThreeWay_Terms = length(threeway_terms)
  ))

@@

@@

# Remove any NULL entries if they exist
models <- models[!sapply(models, is.null)]
if (length(models) == 0) {
  stop("No valid models to pool")
}
```

### Model Fitting Summary

```{r model-summary}
kable(model_progress, caption = "Model fitting progress across imputations")
```

@@

```{r pool-results}
# Use mice::pool for robust pooling with Barnard-Rubin df adjustment
# mice::pool automatically applies Barnard-Rubin (1999) correction

# Ensure models is a proper list
if (!is.list(models)) {
  stop("Models must be a list")
}

@@

# Pool using mice with Barnard-Rubin df
pooled_model <- pool(models)
pooled_summary <- summary(pooled_model)

# Extract results - mice::pool provides df column with Barnard-Rubin adjustment
pooled_results <- pooled_summary |>
  mutate(
    Term = term,
    Estimate = estimate,
    SE = std.error,
    t_stat = statistic,
    df = df,  # Barnard-Rubin adjusted degrees of freedom
    P_Value = p.value,
    # Recalculate CIs using t-distribution with proper df
    CI_Lower = estimate - qt(0.975, df) * std.error,
    CI_Upper = estimate + qt(0.975, df) * std.error,
    OR = exp(estimate),
    OR_CI_Lower = exp(CI_Lower),
    OR_CI_Upper = exp(CI_Upper)
  ) |>
  select(Term, Estimate, SE, t_stat, df, P_Value, CI_Lower, CI_Upper, OR, OR_CI_Lower, OR_CI_Upper)

# Get pooled variance-covariance matrix for later calculations
pooled_vcov <- vcov(pooled_model)

# Ensure proper formatting
rownames(pooled_results) <- NULL

@@

```{r pooling-diagnostics}
# Create diagnostics table - now with df from mice::pool
pool_diag <- data.frame(
  Metric = c("Number of imputations", 
             "Number of parameters",
             "Average degrees of freedom (Barnard-Rubin adjusted)",
             "Minimum degrees of freedom",
             "Maximum degrees of freedom"),
  Value = c(n_imputations,
            nrow(pooled_results),
            round(mean(pooled_results$df), 1),
            round(min(pooled_results$df), 1),
            round(max(pooled_results$df), 1))
)

@@

# Show key terms with their df and p-values
key_terms <- grep("ever_lapse", pooled_results$Term, value = TRUE)[1:min(3, length(grep("ever_lapse", pooled_results$Term, value = TRUE)))]
key_results <- pooled_results[pooled_results$Term %in% key_terms, c("Term", "Estimate", "SE", "df", "P_Value")]
key_results$Term <- substr(key_results$Term, 1, 40)
key_results$Estimate <- round(key_results$Estimate, 3)
key_results$SE <- round(key_results$SE, 3)
key_results$df <- round(key_results$df, 1)
key_results$P_Value <- round(key_results$P_Value, 4)

kable(key_results, caption = "Key lapse-related terms with degrees of freedom")

@@

@@

# Test 1: Three-way interaction
if (length(threeway_terms_test) > 0) {
  hypothesis_3way <- paste0(threeway_terms_test, " = 0")
  pooled_test_3way <- pool_chi_square_tests(models, hypothesis_3way, "Three-way interaction")
}

@@

if (length(lapse_dr_terms) > 0) {
  hypothesis_lapse_dr <- paste0(lapse_dr_terms, " = 0")
  pooled_test_lapse_dr <- pool_chi_square_tests(models, hypothesis_lapse_dr, "Lapse × DR interaction")
}

@@

if (length(lapse_tx_terms) > 0) {
  hypothesis_lapse_tx <- paste0(lapse_tx_terms, " = 0")
  pooled_test_lapse_tx <- pool_chi_square_tests(models, hypothesis_lapse_tx, "Lapse × treatment interaction")
}
```

### Joint Hypothesis Test Results

```{r joint-test-results}
# Compile all test results
joint_tests <- data.frame(
  Test = character(),
  Chi_Square = numeric(),
  df = numeric(),
  P_Value = character(),
  Significant = character(),
  stringsAsFactors = FALSE
)

if (exists("pooled_test_3way")) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × DR × Treatment (3-way)",
    Chi_Square = round(pooled_test_3way$D_bar, 3),
    df = pooled_test_3way$df,
    P_Value = format.pval(pooled_test_3way$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_3way$p_value < 0.05, "Yes", "No")
  ))
}

if (exists("pooled_test_lapse_dr")) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × DR Severity",
    Chi_Square = round(pooled_test_lapse_dr$D_bar, 3),
    df = pooled_test_lapse_dr$df,
    P_Value = format.pval(pooled_test_lapse_dr$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_lapse_dr$p_value < 0.05, "Yes", "No")
  ))
}

if (exists("pooled_test_lapse_tx")) {
  joint_tests <- rbind(joint_tests, data.frame(
    Test = "Lapse × Treatment",
    Chi_Square = round(pooled_test_lapse_tx$D_bar, 3),
    df = pooled_test_lapse_tx$df,
    P_Value = format.pval(pooled_test_lapse_tx$p_value, digits = 4, eps = 0.0001),
    Significant = ifelse(pooled_test_lapse_tx$p_value < 0.05, "Yes", "No")
  ))
}

kable(joint_tests, caption = "Pooled joint hypothesis tests (Li-Raghunathan-Rubin method)")
```

@@

@@
